---
title: "Persona Disposition Seed (n=100)"
format: html
params:
  seed: 42
  donor_topk: 3
  gss_min_year: 2016
  atus_samples: ["at2023", "at2024"]
  acs_source_path: "../../data/acs_pums_sample_n100.csv"
---

## Aim

Attach a **disposition layer** — personality (TIPI Big Five), values (moral traditionalism,
egalitarianism, an opinion-strength battery), religion, party ID, ideology, social trust,
usual work hours, general happiness, and real diary-based wake time — onto the 100-person
ACS base sample built in `acs_pums_national_sample.qmd`. Every field here is either
**directly sampled** (the ACS demographics, carried through unchanged) or **donor-matched**
from a real respondent in ANES, GSS, or ATUS via nearest-neighbor Gower distance on shared
demographics — never invented, never LLM-inferred. **Daylight Saving Time is not seeded
anywhere in this table** — no DST opinion, stance, or values-ranking item is pulled from any
source; DST opinion is meant to emerge later from an LLM reasoning over this seed, then be
validated against real polling. This notebook only builds the persona table; it does not run
a simulation or generate an opinion. Output (`data/persona_seed_dataset_n100.csv`) is
**Stage 1** of the P6a persona pipeline (`docs/adr/0016-census-personas-disposition-pipeline.md`,
`docs/briefs/phase-6a-persona-pipeline.md`): Stage 2, built separately, consumes this exact
file to run the LLM backstory/reflection elaboration — this notebook does not touch `src/`
and never calls an LLM. 1-NN matching only, never k-averaging (averaging across donors would
recreate the stereotype centroid the whole disposition layer exists to avoid).

## Imports

```{r}
#| label: imports
suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(stringr); library(readr); library(tibble)
  library(anesr); library(gssr); library(gower); library(ipumsr); library(tidycensus)
})
```

## Config — edit and re-run downward

```{r}
#| label: config
SEED <- params$seed
DONOR_TOPK <- params$donor_topk          # random pick among the K nearest donors, not always
                                          # the single nearest -- avoids many recipients who
                                          # share a common demographic profile all collapsing
                                          # onto the exact same one donor's disposition.
GSS_MIN_YEAR <- params$gss_min_year      # pool recent GSS waves: the social-trust battery is
                                          # asked of only a rotating subsample each wave, so a
                                          # single year gives too thin a donor pool.
ATUS_SAMPLES <- unlist(params$atus_samples)
ACS_SOURCE_PATH <- params$acs_source_path
ATUS_EXTRACT_XML <- "../../data/raw/atus_00002.xml"  # IPUMS-assigned extract name; see §4 for how this is produced

# ANES year is fixed, not parameterized: TIPI + the moral-traditionalism/egalitarianism
# batteries were fielded in the 2016 Time Series Study specifically (verified empirically
# against anesr::timeseries_2016's variable labels -- not present, or not in this combination,
# in other available waves), so there is no year to sweep here.
```

## Harmonization helpers

Four sources code demographics differently (ACS text labels, ANES/ATUS/GSS numeric codes with
their own scales). Gower matching needs **shared buckets**: every recipient/donor pair is
matched on whatever coarseness both sides can express — a 6-way race/ethnicity bucket for
ACS↔ANES and ACS↔ATUS, but a coarser 3-way bucket for ACS↔GSS since GSS's own `race` variable
only distinguishes White/Black/Other.

```{r}
#| label: harmonization-helpers
bucket_education <- function(label_or_code, source) {
  x <- as.character(label_or_code)
  case_when(
    source == "acs" & str_detect(x, regex("bachelor", ignore_case = TRUE)) ~ "Bachelor's degree",
    source == "acs" & str_detect(x, regex("master|doctorate|professional degree", ignore_case = TRUE)) ~ "Graduate or professional degree",
    source == "acs" & str_detect(x, regex("associate|some college|college credit", ignore_case = TRUE)) ~ "Some college or Associate's",
    source == "acs" & str_detect(x, regex("high school|ged", ignore_case = TRUE)) ~ "High school graduate",
    source == "acs" ~ "Less than high school",

    # ANES V161270 codes (1-16, 90, 95) -- see ANES 2016 codebook
    source == "anes" & x %in% as.character(1:8) ~ "Less than high school",
    source == "anes" & x %in% c("9", "90") ~ "High school graduate",
    source == "anes" & x %in% as.character(10:12) ~ "Some college or Associate's",
    source == "anes" & x == "13" ~ "Bachelor's degree",
    source == "anes" & x %in% as.character(14:16) ~ "Graduate or professional degree",

    # GSS degree (0-4)
    source == "gss" & x == "0" ~ "Less than high school",
    source == "gss" & x == "1" ~ "High school graduate",
    source == "gss" & x == "2" ~ "Some college or Associate's",
    source == "gss" & x == "3" ~ "Bachelor's degree",
    source == "gss" & x == "4" ~ "Graduate or professional degree",

    # ATUS/IPUMS EDUC codes
    source == "atus" & x %in% as.character(10:17) ~ "Less than high school",
    source == "atus" & x %in% c("20", "21") ~ "High school graduate",
    source == "atus" & x %in% c("30", "31", "32") ~ "Some college or Associate's",
    source == "atus" & x == "40" ~ "Bachelor's degree",
    source == "atus" & x %in% c("41", "42", "43") ~ "Graduate or professional degree",
    TRUE ~ NA_character_
  )
}

bucket_marital <- function(label_or_code, source) {
  x <- as.character(label_or_code)
  case_when(
    # order matters: "Never married" also contains the substring "married"
    source == "acs" & str_detect(x, regex("never married", ignore_case = TRUE)) ~ "Never married",
    source == "acs" & str_detect(x, regex("married", ignore_case = TRUE)) ~ "Married",
    source == "acs" & str_detect(x, regex("widow", ignore_case = TRUE)) ~ "Widowed",
    source == "acs" & str_detect(x, regex("divorc", ignore_case = TRUE)) ~ "Divorced",
    source == "acs" & str_detect(x, regex("separat", ignore_case = TRUE)) ~ "Separated",

    source %in% c("anes", "atus") & x %in% c("1", "2") ~ "Married",
    source %in% c("anes", "atus") & x == "3" ~ "Widowed",
    source %in% c("anes", "atus") & x == "4" ~ "Divorced",
    source %in% c("anes", "atus") & x == "5" ~ "Separated",
    source %in% c("anes", "atus") & x == "6" ~ "Never married",

    source == "gss" & x == "1" ~ "Married",
    source == "gss" & x == "2" ~ "Widowed",
    source == "gss" & x == "3" ~ "Divorced",
    source == "gss" & x == "4" ~ "Separated",
    source == "gss" & x == "5" ~ "Never married",
    TRUE ~ NA_character_
  )
}

# 6-way race/ethnicity bucket: ACS <-> ANES, ACS <-> ATUS
bucket_race6_acs <- function(race_ethnicity) {
  case_when(
    race_ethnicity == "White alone" ~ "White",
    race_ethnicity == "Black or African American alone" ~ "Black",
    race_ethnicity %in% c("Asian alone", "Native Hawaiian or Other Pacific Islander alone") ~ "Asian or Pacific Islander",
    race_ethnicity == "American Indian or Alaska Native alone" ~ "American Indian or Alaska Native",
    race_ethnicity == "Hispanic or Latino (any race)" ~ "Hispanic or Latino",
    TRUE ~ "Other or Two or more races"
  )
}
bucket_race6_anes <- function(code) {
  # ANES V161310x is already a combined race+ethnicity summary (Hispanic overrides race),
  # matching the ACS convention above exactly.
  case_when(
    code == "1" ~ "White", code == "2" ~ "Black",
    code == "3" ~ "Asian or Pacific Islander", code == "4" ~ "American Indian or Alaska Native",
    code == "5" ~ "Hispanic or Latino", code == "6" ~ "Other or Two or more races",
    TRUE ~ NA_character_
  )
}
bucket_race6_atus <- function(race_code, hispan_code) {
  # HISPAN: 100 = Not Hispanic, 2xx = specific Hispanic origins (confirmed against the
  # downloaded extract's value distribution, not assumed from the codebook alone).
  is_hisp <- !is.na(hispan_code) & hispan_code != "100"
  case_when(
    is_hisp ~ "Hispanic or Latino",
    race_code == "100" ~ "White",
    race_code == "110" ~ "Black",
    race_code == "120" ~ "American Indian or Alaska Native",
    race_code %in% c("130", "131", "132") ~ "Asian or Pacific Islander",
    TRUE ~ "Other or Two or more races"
  )
}
# 3-way coarse race bucket: ACS <-> GSS (GSS's own ceiling of granularity)
bucket_race3_acs <- function(race_ethnicity) {
  case_when(
    race_ethnicity == "White alone" ~ "White",
    race_ethnicity == "Black or African American alone" ~ "Black",
    TRUE ~ "Other"
  )
}
bucket_race3_gss <- function(code) {
  case_when(code == "1" ~ "White", code == "2" ~ "Black", code == "3" ~ "Other", TRUE ~ NA_character_)
}

# gower::gower_topn mutates its input columns IN PLACE at the C level for character
# columns (confirmed by bisection: after a gower_topn() call, the CALLER's original
# tibble had cells overwritten with values copied from OTHER ROWS of that same tibble,
# even though the arguments were built via `select()` moments earlier -- R's normal
# copy-on-write does not protect against this). Two independent fixes are required:
# (1) convert character matching columns to factors before calling gower_topn, which
# avoids the mutation; (2) those factors must share IDENTICAL levels between x and y,
# or gower_topn raises "Levels ... do not match".
gower_safe_pair <- function(x_df, y_df) {
  char_cols <- names(x_df)[vapply(x_df, is.character, logical(1))]
  for (col in char_cols) {
    lv <- union(unique(x_df[[col]]), unique(y_df[[col]]))
    x_df[[col]] <- factor(x_df[[col]], levels = lv)
    y_df[[col]] <- factor(y_df[[col]], levels = lv)
  }
  list(x = as.data.frame(x_df), y = as.data.frame(y_df))
}

#' Nearest-donor picker: random choice among the DONOR_TOPK nearest ties, seeded per
#' recipient row so the whole pipeline is deterministic end-to-end under SEED.
pick_donor <- function(recipient_keys, donor_keys, seed, k = DONOR_TOPK) {
  safe <- gower_safe_pair(recipient_keys, donor_keys)
  topn <- gower_topn(safe$x, safe$y, n = k)
  n_recip <- nrow(safe$x)
  chosen <- integer(n_recip); chosen_dist <- numeric(n_recip)
  for (i in seq_len(n_recip)) {
    set.seed(seed * 100000 + i)
    cand <- topn$index[, i]; dist <- topn$distance[, i]
    cand <- cand[!is.na(cand)]; dist <- dist[!is.na(dist)]
    pick <- sample(seq_along(cand), 1)
    chosen[i] <- cand[pick]; chosen_dist[i] <- dist[pick]
  }
  list(donor_row = chosen, distance = chosen_dist)
}
```

## 1. Recipients — the ACS base sample

Reuses the already-built, already-validated national sample from
`acs_pums_national_sample.qmd` (`data/acs_pums_sample_n100.csv`) rather than re-pulling ACS
PUMS here — that pull takes ~10 minutes against the Census API and this notebook answers a
different question (disposition matching, not demographic sampling).

```{r}
#| label: recipients
acs <- read_csv(ACS_SOURCE_PATH, show_col_types = FALSE)
cat("ACS base rows:", nrow(acs), "\n")

recipients <- acs %>%
  transmute(
    person_id, age, sex, region,
    race6 = bucket_race6_acs(race_ethnicity),
    race3 = bucket_race3_acs(race_ethnicity),
    education_bucket = bucket_education(education, "acs"),
    marital_bucket = bucket_marital(marital_status, "acs")
  )
stopifnot(nrow(recipients) == 100, sum(is.na(recipients$education_bucket)) == 0)
cat("recipients built:\n")
print(head(recipients, 3))
```

## 2. ANES donor pool — personality, values, religion, party, ideology

`anesr::timeseries_2016` (the 2016 ANES Time Series Study, bundled directly by the package —
no gated ICPSR download needed). Variable codes verified empirically against the bundled
labels before use, not assumed from memory:

- **TIPI** (Ten-Item Personality Inventory, Gosling/Rentfrow/Swann): `V162333`–`V162342`,
  reverse-scored per the standard TIPI key into five 1–7 trait scores.
- **Moral traditionalism** (Feldman 4-item scale): `V162207`–`V162210`.
- **Egalitarianism** (Feldman–Steenbergen 4-item scale): `V162243`–`V162246`.
- **Opinion strength**: `V162248`–`V162250` — ANES has no Cacioppo–Petty Need-for-Cognition
  battery; this is the closest real ANES battery (a "need to evaluate"/opinionatedness
  measure) and is labeled honestly as `opinion_strength`, not mislabeled as NFC.
- **Party ID** (`V161158x`, 7-pt summary), **ideology** (`V161126`, 7-pt summary).
- **Religion**: major-group summary (`V161265x`, 8 categories), importance (`V161241`),
  attendance (`V161244` gate + `V161245`).

```{r}
#| label: anes-pool
data(timeseries_2016)
anes_raw <- timeseries_2016

anes <- anes_raw %>%
  transmute(anes_respondent_id = V160001, across(
    c(V161267, V161342, V161310x, V161270, V161268, V163003,
      V162333, V162334, V162335, V162336, V162337, V162338, V162339, V162340, V162341, V162342,
      V162207, V162208, V162209, V162210,
      V162243, V162244, V162245, V162246,
      V162248, V162249, V162250,
      V161158x, V161126, V161265x, V161241, V161244, V161245),
    as.numeric
  ))

anes <- anes %>%
  mutate(
    age = ifelse(V161267 > 0, V161267, NA),
    sex = case_when(V161342 == 1 ~ "Male", V161342 == 2 ~ "Female", TRUE ~ NA_character_),
    race6 = bucket_race6_anes(as.character(ifelse(V161310x > 0, V161310x, NA))),
    education_bucket = bucket_education(ifelse(V161270 > 0, V161270, NA), "anes"),
    marital_bucket = bucket_marital(ifelse(V161268 > 0, V161268, NA), "anes"),
    region = case_when(V163003 == 1 ~ "Northeast", V163003 == 2 ~ "Midwest",
                        V163003 == 3 ~ "South", V163003 == 4 ~ "West", TRUE ~ NA_character_),
    # TIPI pairs: (typical item + reverse-keyed item) / 2, on the original 1-7 scale
    tipi_extraversion = (ifelse(V162333 > 0, V162333, NA) + (8 - ifelse(V162338 > 0, V162338, NA))) / 2,
    tipi_agreeableness = ((8 - ifelse(V162334 > 0, V162334, NA)) + ifelse(V162339 > 0, V162339, NA)) / 2,
    tipi_conscientiousness = (ifelse(V162335 > 0, V162335, NA) + (8 - ifelse(V162340 > 0, V162340, NA))) / 2,
    tipi_emotional_stability = ((8 - ifelse(V162336 > 0, V162336, NA)) + ifelse(V162341 > 0, V162341, NA)) / 2,
    tipi_openness = (ifelse(V162337 > 0, V162337, NA) + (8 - ifelse(V162342 > 0, V162342, NA))) / 2,
    moral_traditionalism = ((6 - ifelse(V162207 > 0, V162207, NA)) + ifelse(V162208 > 0, V162208, NA) +
                             (6 - ifelse(V162209 > 0, V162209, NA)) + ifelse(V162210 > 0, V162210, NA)) / 4,
    egalitarianism = (ifelse(V162243 > 0, V162243, NA) + (6 - ifelse(V162244 > 0, V162244, NA)) +
                       (6 - ifelse(V162245 > 0, V162245, NA)) + ifelse(V162246 > 0, V162246, NA)) / 4,
    opinion_strength = (ifelse(V162248 > 0, V162248, NA) + ifelse(V162249 > 0, V162249, NA) + ifelse(V162250 > 0, V162250, NA)) / 3,
    party_id_7pt = ifelse(V161158x > 0, V161158x, NA),
    ideology_7pt = ifelse(V161126 %in% 1:7, V161126, NA),
    religion_group = case_when(
      V161265x == 1 ~ "Mainline Protestant", V161265x == 2 ~ "Evangelical Protestant",
      V161265x == 3 ~ "Black Protestant", V161265x == 4 ~ "Roman Catholic",
      V161265x == 5 ~ "Undifferentiated Christian", V161265x == 6 ~ "Jewish",
      V161265x == 7 ~ "Other religion", V161265x == 8 ~ "Not religious", TRUE ~ NA_character_
    ),
    religion_importance = case_when(V161241 == 1 ~ "Important", V161241 == 2 ~ "Not important", TRUE ~ NA_character_),
    religion_attendance = case_when(
      V161244 == 2 ~ "Never",
      V161245 == 1 ~ "Every week", V161245 == 2 ~ "Almost every week",
      V161245 == 3 ~ "Once or twice a month", V161245 == 4 ~ "A few times a year",
      V161245 == 5 ~ "Never", TRUE ~ NA_character_
    )
  )

MATCHING_COLS <- c("age", "sex", "race6", "education_bucket", "marital_bucket", "region")
CONTENT_COLS_ANES <- c("tipi_extraversion", "tipi_agreeableness", "tipi_conscientiousness",
                        "tipi_emotional_stability", "tipi_openness", "moral_traditionalism",
                        "egalitarianism", "opinion_strength", "party_id_7pt", "ideology_7pt",
                        "religion_group", "religion_importance", "religion_attendance")

anes_pool <- anes %>% select(all_of(c("anes_respondent_id", MATCHING_COLS, CONTENT_COLS_ANES))) %>%
  filter(if_all(everything(), ~ !is.na(.)))
cat("ANES donor pool (complete cases):", nrow(anes_pool), "of", nrow(anes), "\n")

anes_match <- pick_donor(recipients %>% select(all_of(MATCHING_COLS)),
                          anes_pool %>% select(all_of(MATCHING_COLS)), seed = 1001)
cat("ANES match distance  min/median/max:",
    round(min(anes_match$distance), 3), round(median(anes_match$distance), 3), round(max(anes_match$distance), 3), "\n")
cat("ANES donor reuse (max times one donor picked):", max(table(anes_match$donor_row)), "\n")
```

## 3. GSS donor pool — social trust, work hours, happiness (supplementary)

Kept deliberately narrow — only content ANES/ATUS don't cover. GSS's own `race` variable is
coarse (White/Black/Other), so matching here uses `race3`, not `race6`. The social-trust
battery is asked of only a rotating subsample per wave, so the 5 most recent waves are pooled
for an adequate donor pool.

```{r}
#| label: gss-pool
data(gss_all)
gss_recent <- gss_all %>% filter(year >= GSS_MIN_YEAR)

gss <- gss_recent %>%
  transmute(
    gss_respondent_id = paste0("gss_", as.numeric(year), "_", as.numeric(id)),
    age = suppressWarnings(as.numeric(age)),
    sex = case_when(as.numeric(sex) == 1 ~ "Male", as.numeric(sex) == 2 ~ "Female", TRUE ~ NA_character_),
    race3 = bucket_race3_gss(as.character(as.numeric(race))),
    education_bucket = bucket_education(as.numeric(degree), "gss"),
    marital_bucket = bucket_marital(as.numeric(marital), "gss"),
    region = case_when(as.numeric(region) == 1 ~ "Northeast", as.numeric(region) == 2 ~ "Midwest",
                        as.numeric(region) == 3 ~ "South", as.numeric(region) == 4 ~ "West", TRUE ~ NA_character_),
    social_trust = case_when(as.numeric(trust) == 1 ~ "Can trust people",
                              as.numeric(trust) == 2 ~ "Can't be too careful",
                              as.numeric(trust) == 3 ~ "Depends", TRUE ~ NA_character_),
    usual_hours_worked = suppressWarnings(as.numeric(hrs1)),
    general_happiness = case_when(as.numeric(happy) == 1 ~ "Very happy", as.numeric(happy) == 2 ~ "Pretty happy",
                                   as.numeric(happy) == 3 ~ "Not too happy", TRUE ~ NA_character_)
  )

MATCHING_COLS_GSS <- c("age", "sex", "race3", "education_bucket", "marital_bucket", "region")
CONTENT_COLS_GSS <- c("social_trust", "usual_hours_worked", "general_happiness")

gss_pool <- gss %>% select(all_of(c("gss_respondent_id", MATCHING_COLS_GSS, CONTENT_COLS_GSS))) %>%
  filter(if_all(everything(), ~ !is.na(.)))
cat("GSS donor pool (complete cases, waves >=", GSS_MIN_YEAR, "pooled):", nrow(gss_pool), "\n")

gss_match <- pick_donor(recipients %>% select(all_of(MATCHING_COLS_GSS)),
                         gss_pool %>% select(all_of(MATCHING_COLS_GSS)), seed = 2002)
cat("GSS match distance  min/median/max:",
    round(min(gss_match$distance), 3), round(median(gss_match$distance), 3), round(max(gss_match$distance), 3), "\n")
cat("GSS donor reuse (max):", max(table(gss_match$donor_row)), "\n")
```

## 4. ATUS donor pool — real diary-based wake time

Pulled via **IPUMS Time Use** (`ipumsr`, `collection = "atus"`) rather than BLS's own site —
BLS's public-use-file pages return HTTP 403 to every automated fetch from this environment
(bot-protection, not a login wall), and its NBER mirror does too. IPUMS Time Use needs a free
account + API key (`IPUMS_TIMEUSE_API_KEY` in `.env`), which is a `.env`-documented
credential like `CENSUS_API`, not a manual per-run download.

Wake time is **derived from real activity-episode diary data**, not a ready-made variable:
each respondent's day is a sequence of activities with `START`/`STOP` clock times; the sleep
activity code is `10101` (confirmed against the extract's own value labels, family
`101xx`). Wake time = the `STOP` time of the diary day's first activity, *if* that first
activity is sleep (the ATUS diary always starts at 04:00, so a respondent already awake at
04:00 has no observed wake time this diary day — coded `NA`, not guessed). Clock times before
04:00 belong to the *end* of the diary day, not the start, so a diary-relative minute offset
is used for "first," not a raw string/clock-time sort.

```{r}
#| label: atus-extract
#| eval: false
# Run once to build the extract; re-running is safe (IPUMS extracts are idempotent to
# re-request, just re-download the same XML if it already exists locally).
ipums_key_line <- readLines("../../.env", warn = FALSE)
ipums_key_line <- ipums_key_line[grepl("^IPUMS_TIMEUSE_API_KEY=", ipums_key_line)]
ipums_key <- trimws(gsub("[\"']", "", sub("^IPUMS_TIMEUSE_API_KEY=", "", ipums_key_line[1])))
set_ipums_api_key(ipums_key, save = FALSE)

ext <- define_extract_micro(
  collection = "atus",
  description = "polis disposition seed - wake time donor pool",
  samples = ATUS_SAMPLES,
  variables = c("AGE", "SEX", "RACE", "HISPAN", "EDUC", "MARST", "STATEFIP",
                "ACTIVITY", "START", "STOP", "WT06"),
  data_structure = "rectangular",
  rectangular_on = "A"   # activity-level rows -- needed for START/STOP per episode
)
ready <- wait_for_extract(submit_extract(ext), timeout_seconds = 300)
dir.create("../../data/raw", showWarnings = FALSE)
download_extract(ready, download_dir = "../../data/raw", overwrite = TRUE)
# IPUMS assigns the extract number itself (e.g. atus_00003.xml on a fresh account) --
# the DDI's internal fileName reference is tied to its paired .dat.gz, so don't rename
# either file after download. Update ATUS_EXTRACT_XML above to match whatever name this
# produces before re-running §4 downstream.
```

```{r}
#| label: atus-pool
# state FIPS -> region, reusing the same Census Bureau 4-region assignment as the ACS
# notebook, joined through tidycensus's verified FIPS/abbreviation crosswalk rather than
# re-typing FIPS codes by hand.
REGION_LOOKUP_ABBR <- tribble(
  ~state, ~region,
  "CT", "Northeast", "ME", "Northeast", "MA", "Northeast", "NH", "Northeast",
  "RI", "Northeast", "VT", "Northeast", "NJ", "Northeast", "NY", "Northeast", "PA", "Northeast",
  "IL", "Midwest", "IN", "Midwest", "MI", "Midwest", "OH", "Midwest", "WI", "Midwest",
  "IA", "Midwest", "KS", "Midwest", "MN", "Midwest", "MO", "Midwest", "NE", "Midwest",
  "ND", "Midwest", "SD", "Midwest",
  "DE", "South", "FL", "South", "GA", "South", "MD", "South", "NC", "South", "SC", "South",
  "VA", "South", "DC", "South", "WV", "South", "AL", "South", "KY", "South", "MS", "South",
  "TN", "South", "AR", "South", "LA", "South", "OK", "South", "TX", "South",
  "AZ", "West", "CO", "West", "ID", "West", "MT", "West", "NV", "West", "NM", "West",
  "UT", "West", "WY", "West", "AK", "West", "CA", "West", "HI", "West", "OR", "West", "WA", "West"
)
FIPS_TO_REGION <- tidycensus::fips_codes %>%
  distinct(state, state_code) %>%
  inner_join(REGION_LOOKUP_ABBR, by = "state") %>%
  transmute(STATEFIP = as.numeric(state_code), region = region)

to_diary_minutes <- function(hms) {
  parts <- str_split_fixed(hms, ":", 3)
  h <- as.numeric(parts[, 1]); m <- as.numeric(parts[, 2])
  ifelse(h < 4, (h + 24) * 60 + m, h * 60 + m)
}

atus_ddi <- read_ipums_ddi(ATUS_EXTRACT_XML)
atus_raw <- read_ipums_micro(atus_ddi)
cat("ATUS activity rows:", nrow(atus_raw), " persons:", n_distinct(atus_raw$CASEID), "\n")

atus_act <- atus_raw %>%
  mutate(across(c(AGE, SEX, RACE, HISPAN, EDUC, MARST, STATEFIP, ACTIVITY), as.numeric)) %>%
  mutate(is_sleep = ACTIVITY %in% c(10100, 10101, 10102, 10199),
         start_diary_min = to_diary_minutes(START))

first_activity <- atus_act %>%
  group_by(CASEID) %>%
  slice_min(start_diary_min, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(CASEID, wake_time = ifelse(is_sleep, STOP, NA_character_))

atus_demo <- atus_act %>%
  group_by(CASEID) %>%
  summarise(AGE = first(AGE), SEX = first(SEX), RACE = first(RACE), HISPAN = first(HISPAN),
            EDUC = first(EDUC), MARST = first(MARST), STATEFIP = first(STATEFIP), .groups = "drop")

atus <- atus_demo %>%
  left_join(first_activity, by = "CASEID") %>%
  left_join(FIPS_TO_REGION, by = "STATEFIP") %>%
  mutate(
    age = AGE,
    sex = case_when(SEX == 1 ~ "Male", SEX == 2 ~ "Female", TRUE ~ NA_character_),
    race6 = bucket_race6_atus(as.character(RACE), as.character(HISPAN)),
    education_bucket = bucket_education(EDUC, "atus"),
    marital_bucket = bucket_marital(MARST, "atus")
  )

atus_pool <- atus %>% transmute(atus_respondent_id = as.character(CASEID), across(all_of(c(MATCHING_COLS, "wake_time")))) %>%
  filter(if_all(everything(), ~ !is.na(.)))
cat("ATUS donor pool (complete cases, wake time observed):", nrow(atus_pool), "\n")

atus_match <- pick_donor(recipients %>% select(all_of(MATCHING_COLS)),
                          atus_pool %>% select(all_of(MATCHING_COLS)), seed = 3003)
cat("ATUS match distance  min/median/max:",
    round(min(atus_match$distance), 3), round(median(atus_match$distance), 3), round(max(atus_match$distance), 3), "\n")
cat("ATUS donor reuse (max):", max(table(atus_match$donor_row)), "\n")
```

## 5. Assemble and export

Every donor-matched column is suffixed by source (`_anes` / `_gss` / `_atus`) so provenance
is self-documenting in the column names, not only in this prose. Per the P6a joint-dataset
schema contract (`docs/briefs/phase-6a-persona-pipeline.md`), each source also carries its
real **donor respondent id** alongside its match distance — R29 provenance, so a persona's
disposition traces back to an actual survey respondent, not just "some ANES row."

```{r}
#| label: assemble
final_table <- acs %>%
  select(person_id, state_abbr, state_name, region, puma, sex, age, race_ethnicity,
         hispanic_origin, education, occupation, industry, marital_status,
         household_relationship, presence_of_children, family_size,
         gave_birth_past_year, person_weight) %>%
  bind_cols(
    anes_pool[anes_match$donor_row, c("anes_respondent_id", CONTENT_COLS_ANES)] %>% rename_with(~ paste0(., "_anes"), -anes_respondent_id),
    tibble(anes_match_distance = anes_match$distance),
    gss_pool[gss_match$donor_row, c("gss_respondent_id", CONTENT_COLS_GSS)] %>% rename_with(~ paste0(., "_gss"), -gss_respondent_id),
    tibble(gss_match_distance = gss_match$distance),
    atus_pool[atus_match$donor_row, c("atus_respondent_id", "wake_time")] %>% rename(wake_time_atus = wake_time),
    tibble(atus_match_distance = atus_match$distance)
  )

stopifnot(nrow(final_table) == 100)
cat("final_table:", nrow(final_table), "rows x", ncol(final_table), "cols\n")

# filename matches the P6a brief's joint-dataset schema contract exactly, so the
# Stage-2 persona pipeline (built separately, against a synthetic fixture) can pick
# this up as its real input without a path change.
write_csv(final_table, "../../data/persona_seed_dataset_n100.csv")
cat("wrote data/persona_seed_dataset_n100.csv\n")
```

## 6. Provenance and match-quality summary

```{r}
#| label: provenance
provenance <- tribble(
  ~field_group, ~columns, ~source, ~method,
  "Demographics (18 cols)", "state_abbr..person_weight", "ACS PUMS (2024 1-yr)", "Direct-sampled (weighted draw)",
  "Personality/values/religion (13 cols + id + distance)", "tipi_*..religion_attendance_anes, anes_respondent_id, anes_match_distance", "ANES 2016 Time Series", "Donor-matched (Gower 1-NN among top-3 ties, never averaged)",
  "Social trust/hours/happiness (3 cols + id + distance)", "social_trust_gss..general_happiness_gss, gss_respondent_id, gss_match_distance", "GSS 2016-2024 pooled", "Donor-matched (Gower 1-NN among top-3 ties, never averaged)",
  "Wake time (1 col + id + distance)", "wake_time_atus, atus_respondent_id, atus_match_distance", "ATUS via IPUMS Time Use 2023-2024", "Donor-matched (Gower 1-NN among top-3 ties, never averaged)"
)
print(provenance)

cat("\nDonor pool sizes: ANES", nrow(anes_pool), " GSS", nrow(gss_pool), " ATUS", nrow(atus_pool), "\n")
cat("Max donor reuse:  ANES", max(table(anes_match$donor_row)),
    " GSS", max(table(gss_match$donor_row)),
    " ATUS", max(table(atus_match$donor_row)), "\n")
cat("\nNo Daylight Saving Time item of any kind appears in this table -- confirmed by column list above.\n")
```
